Think deep for reef resilience

### 2025-07-15 paper/paper_draft.qmd
Add caveat about non- bias corrected DHW data
Edit intro for flow, conciseness
Remove reference to nutrients, cyclones etc, that aren't explored/relevant to the study.

Climate change is impacting species survival, persistence and ecosystem functions globally,
including coral reefs. Coral reefs are vulnerable to increasing magnitude and variation in
environmental stressors such as ocean temperature, acidification and nutrient loading
[@foster2021; @gattuso2014]. Mass bleaching and mortality events linked to climate change
can cause reef-building coral population declines, possibly leading to long-term and
large-scale coral-macroalgae regime shifts [@arif2022; @wild2011]. Climate impacts on coral
reefs are, and will continue to, decrease the ability of coral reefs to provide ecological
functions such as benthic habitat and nutrient cycling [@forster2021; @williams2019] as well
as ecosystem services such as food, protection, tourism and Indigenous cultural value
[@eddy2021; @stoeckl2021]. We aim to assess the impacts of possible climate futures and
identify the characteristics of resilient reefs using the latest generation of larval
connectivity research with scenario-based population modelling for the Great Barrier Reef
ecosystem.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of
thousands of individual reefs. They are interconnected through coral and fish larval
dispersal patterns mediated by hydrodynamic, spatial, and ecological processes
[@grimaldi2022]. Current modelling projects severe declines in coral cover under projections
of Degree Heating Weeks [@bozec]. These coral cover declines will be compounded by ocean
acidification, intensifying tropical cyclones, and crown-of-thorns starfish outbreaks
[@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. Ominously, the
2023-24 Summer mass bleaching event may be signalling a shift towards this reef future, with
some level of bleaching recorded on 79% of surveyed reefs [@cantin2024].

Connectivity is said to play a crucial role in Great Barrier Reef resilience by enabling
recolonization after disturbances and facilitating adaptation to climate pressures
[@angeler2023; @elmhirst2009; @hock2017]. This connectivity, however, creates both strengths
and vulnerabilities within the system. Larval source populations (those exporting more
larvae than they import) are essential for system-wide resilience, as they provide larvae
that enable recovery at distant locations after disturbances [@hock2017; @thomas2015].
Individually, however, they are more vulnerable to extreme disturbances as they rely more
heavily on local recruitment to support recovery. In contrast, sink populations can draw on
external larval inputs from multiple and diverse sources [@ani2024; @hock2017]. When
disturbances impact major larval source populations, the cascading effects extend throughout
the network. Increased adult mortality reduces larval export in subsequent years, slowing
recovery at the connected sink locations that depend on this external supply
[@figueiredo2022; @hock2017; @munday2009]. Future climate change will further complicate
these dynamics through population fragmentation, increased larval mortality, and altered
hydrodynamics, potentially disrupting the connectivity patterns that currently maintain reef
resilience.

Beyond connectivity, coral resilience to climate change is determined by the complex
interplay of environmental and biological factors. Environmental abiotic factors influence
how reefs will experience climate change pressures, including the spatio-temporal patterns
of thermal stress itself, and indirect factors that influence the stress experienced by
corals. Environmental factors such as tropical cyclones and wind driven mixing [@lin2023;
@mcclanahan2005]; currents [@grimaldi2022] and depth [@álvarez-noriega2018; @baird2018;
@smith2014] can decrease the magnitude of thermal stress corals experience. Conversely, high
light intensity and nutrient levels in the benthic zone can increase coral bleaching
[@decarlo2020; @higuchi2015].

Individual coral resilience depends on multiple biological factors beyond connectivity.
Coral size [@brandt2009], taxa [@álvarez-noriega2018; @mcclanahan2005], and acclimation and
inherited thermal tolerance [@humanes2022; @lachs2024] all influence the outcome of high
coral heat stress events. Importantly, corals have demonstrated adaptive capacity to develop
thermal tolerance levels capable of delaying the onset of widespread bleaching under low
emissions scenarios [@bairos-novak2021; @humanes2022]. This adaptive potential represents a
crucial mechanism for reef persistence under moderate climate change, working in concert
with the environmental factors that modulate thermal stress exposure.

While these environmental and biological factors determine whether individual corals survive
thermal stress events, the long-term viability of coral reefs depends not only on coral
survival but also on the maintenance of reef structure through positive carbonate budgets.
Scleractinian corals are primary habitat structures on coral reefs, and the accumulation and
erosion of coral reef carbonate frameworks is controlled by a range of processes including
coral accumulation [@perry2015], bioerosion , physical damage generating coral rubble
[@kenyon], and calcification and CaCO₃ dissolution rates [@eyre2014; @eyre2018]. The balance
of these processes and occurrence of net carbonate framework production (referred to as a
positive carbonate budget) are influenced by factors such as coral cover and complexity
[@perry2012; @desbiens], rates of physical environmental processes, and ocean acidity
[@perry2015; @eyre2018]. Total live coral cover on a reef highly influences the maintenance
of net positive carbonate budgets, with estimates ranging from 10 to 20% [@perry2015] of a
reef's total area required to maintain positive budgets. For mid and outer shelf reefs on
the Great Barrier Reef, this threshold is reportedly 17% live coral cover [@desbiens].

Incorporating the latest advancements in whole ecosystem larval connectivity modeling and
coral thermal response and adaptation research into coral population modeling allows an
investigation into how abiotic and biotic factors interact to influence coral climate change
resilience and the maintenance of positive carbonate budgets. Improvements in coral reef
climate response modeling, including sampling from a range of current projected climate
futures, enable timeseries analyses of uncertain climate change and ecological scenarios
[@bode2017; @chollett2022; @donner2018; @klein2024; @uusitalo2015].

It has previously been suggested that highly connected source reefs that are resilient to
bleaching and CoTS outbreaks are likely to play a critical role in supporting future reef
recovery across the entire GBR [@hock2017]. Integrated modelling incorporating data from the
latest advancements in coral population modelling [@ribeirodealmeida2024], whole of
ecosystem larval connectivity \[**citation**\], and current projections of Degree Heating
Weeks [@dixon2022; @dixon] indicate that depth, rather than connectivity, plays a more
critical role in reef resilience under projected conditions. It is further demonstrated that
deeper reefs support maintenance of positive carbonate budget under all climate projections.

@@

There are multiple caveats in how coral ecosystem dynamics are modelled in this
investigation. We model coral colonies as objects with circular diameter values and track
these diameters over timesteps through growth. Due to the availability of quality future
projected data we do not consider the impacts of waves, cyclones and currents on coral
colony survival and bleaching (insert citations for waves affecting survival and bleaching).
We do not consider the impacts of Crown of Thorns Starfish (CoTS) on coral survival and reef
resilience (insert citation for cots affecting corals). Furthermore, there are multiple
caveats to acknowledge in the data sources used in this investigation. Uncertainty in
connectivity modelling outputs, satellite-derived bathymetry data and future thermal stress
projections are likely to influence the results obtained in this investigation. To limit the
influence of bathymetry data on results, reefs with less than 5% data coverage are removed
from analyses. We investigate a range of future climate responses and thermal projections to
account for a range of possible futures, and our results suggest the factors that influence
reef resilience are consistent across these scenarios.

>>>>

Climate change is impacting species survival, persistence and ecosystem functions globally,
including coral reefs. Coral reefs are vulnerable to increasing magnitude and variation in
environmental stressors such as ocean temperature, acidification and nutrient loading
[@foster2021; @gattuso2014]. Mass bleaching and mortality events linked to climate change
can cause reef-building coral population declines, possibly leading to long-term and
large-scale coral-macroalgae regime shifts [@arif2022; @wild2011]. Climate impacts on coral
reefs are, and will continue to, decrease the ability of coral reefs to provide ecological
functions such as benthic habitat and nutrient cycling [@forster2021; @williams2019] as well
as ecosystem services such as food, protection, tourism and Indigenous cultural value
[@eddy2021; @stoeckl2021].

Understanding the characteristics of reefs that will persist under projected climate
conditions requires examining reef resilience across multiple scales and processes.
Structural resilience - the capacity to maintain coral populations and avoid ecosystem
regime shifts - depends on three interconnected mechanisms. First, the ability of individual
corals to survival thermal stress. Second, the connectivity networks that enable
recolonisation after disturbances. Third, the maintenance of carbonate framework that
provides reef structure. We aim to assess how these mechanisms interact under current
projected climate conditions to identify the characteristics of structurally resilient reefs through scenario-based coral population modelling for the Great Barrier Reef ecosystem.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of
thousands of individual reefs. They are interconnected through coral and fish larval
dispersal patterns mediated by hydrodynamic, spatial, and ecological processes
[@grimaldi2022]. The GBR faces severe projected coral cover declines under climate change [@bozec]. These coral cover declines will be compounded by ocean acidification, intensifying tropical cyclones, and crown-of-thorns starfish outbreaks
[@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. The 2023-24
Summer mass bleaching event may be signalling a shift towards this reef future, with some
level of bleaching recorded on 79% of surveyed reefs [@cantin2024].

At the individual coral scale, survival during thermal stress events depends on both
environmental and biological factors. Environmental factors can either amplify or reduce
thermal stress exposure. Depth has been suggested as one crucial factor that can provide
a refuge from thermal stress [@álvarez-noriega2018; @baird2018; @smith2014]. Biological
factors including coral size [@brandt2009], taxa [@álvarez-noriega2018; @mcclanahan2005],
and thermal tolerance through acclimation and inheritance [@humanes2022; @lachs2024] all
influence survival outcomes. Importantly, corals demonstrate adaptive capacity to develop
thermal tolerance that could delay widespread bleaching under moderate emissions scenarios
[@bairos-novak2021; @humanes2022], representing a crucial mechanism for reef persistence.

Beyond individual survival, reef-scale resilience depends on connectivity networks that
enable recolonization after disturbances [@angeler2023; @elmhirst2009; @hock2017]. However,
connectivity creates both strengths and vulnerabilities. Larval source populations that
export more larvae than they import are essential for system-wide recovery, providing larvae
that enable recolonization at distant locations [@hock2017; @thomas2015]. Yet these source
reefs are paradoxically more vulnerable individually, relying heavily on local recruitment
for their own recovery. When disturbances impact major source populations, cascading effects
reduce larval export and slow recovery throughout the connected network [@figueiredo2022;
@hock2017; @munday2009].

Long-term reef viability requires not only coral survival and recolonization, but also
maintenance of the physical reef structure through positive carbonate budgets. Scleractinian
corals build the primary habitat framework, but reef persistence depends on the balance
between carbonate accumulation [@perry2015] and loss through bioerosion, physical damage
[@kenyon], and dissolution [@eyre2014; @eyre2018]. This balance is influenced by coral cover
and complexity [@perry2012; @desbiens], environmental processes, and ocean chemistry
[@perry2015; @eyre2018]. For Great Barrier Reef mid and outer shelf reefs, maintaining
positive carbonate budgets requires approximately 17% live coral cover [@desbiens].

Recent advances in whole-ecosystem larval connectivity modelling and coral thermal response
research enable integrated investigation of how these multiple factors interact to determine
climate resilience. By incorporating the latest coral population models
[@ribeirodealmeida2024], ecosystem-wide connectivity data, and updated climate projections
[@dixon2022], we can examine which factors most strongly influence reef persistence under
different scenarios. While connectivity has been emphasized as critical for GBR resilience
[@hock2017], our integrated modelling reveals that environmental factors - particularly
depth - may play a more fundamental role in determining which reefs maintain viable coral
populations and positive carbonate budgets under projected climate conditions.

@@

There are multiple caveats in how coral ecosystem dynamics are modelled in this
investigation. We model coral colonies as objects with circular diameter values and track
these diameters over timesteps through growth. Due to the availability of quality future
projected data we do not consider the impacts of waves, cyclones and currents on coral
colony survival and bleaching (insert citations for waves affecting survival and bleaching).
We do not consider the impacts of Crown of Thorns Starfish (CoTS) on coral survival and reef
resilience (insert citation for cots affecting corals). Furthermore, there are multiple
caveats to acknowledge in the data sources used in this investigation. Uncertainty in
connectivity modelling outputs, satellite-derived bathymetry data and future thermal stress
projections are likely to influence the results obtained in this investigation. To limit the
influence of bathymetry data on results, reefs with less than 5% data coverage are removed
from analyses. The DHW data used in this investigation have not been bias corrected for the 
Great Barrier Reef region and is part of a wider Australian dataset [@dixon2022] developed 
using statistical downscaling. We investigate a range of future climate responses and 
thermal projections to account for a range of possible futures, and our results suggest 
the factors that influence reef resilience are consistent across these scenarios.
